Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-20 doc/thesis/conclusion.tex

This thesis developed \software{netabc}, a computer program implementing a
statistical inference method for contact network parameters from viral
phylogenetic data. Due to the rapid advancement of nucleotide sequencing
technology, viral sequence data data have become increasingly feasible to
collect on a population level. Through phylodynamic methods, these data offer a
window into epidemiological processes which would otherwise be virtually
impossible to study on a realistic scale. Although phylodynamic methods have
been developed to fit a wide variety of epidemiological models to phylogenetic
data~\autocite{volz2012complex, rasmussen2014phylodynamic}, our method is the
first (to our knowledge) which can fit models that do not assume panmixia. We
believe this capability will be of broad interest to the molecular evolution
and epidemiology community, as it widens the field of epidemiological
parameters which may be investigated through viral sequence data. In addition,
the characterization of local contact networks could be valuable from a public
health perspective, such as for investigating optimal vaccination strategies.
This information could assist in curtailing current epimemics, as well as
preventing future epidemics of different diseases over the same contact
network.

Contact network structure has a substantial impact on epidemic trajectory
\autocite{barthelemy2005dynamical, volz2007susceptible, volz2008sir}, and a few
methods have been developed to estimate network parameters from epidemiological
data \autocite{volz2007susceptible, britton2002bayesian,
groendyke2011bayesian}. It is known that contact network structure can have a
substantial impact on transmission tree shape \autocite{o2010contact,
robinson2013dynamics, leventhal2012inferring, colijn2014phylogenetic,
villandre2016assessment}.

>>>>

Due to the rapid advancement of nucleotide sequencing technology, viral
sequence data data have become increasingly feasible to collect on a population
level. Through phylodynamic methods, these data offer a window into
epidemiological processes which would otherwise be virtually impossible to
study on a realistic scale.

This thesis developed \software{netabc}, a computer program implementing a
statistical inference method for contact network parameters from viral
phylogenetic data. \software{Netabc} brings together the areas of viral
phylodynamics and network epidemiology, which have only intersected in a very
limited fashion thus far~\autocite{welch2011statistical}. The use of
kernel-\gls{ABC}, a likelihood-free method, makes it possible to fit network
models to phylogenies without calculating intractible likelihoods.

Although phylodynamic methods have been developed to fit a wide variety of
epidemiological models to phylogenetic data~\autocite{volz2012complex,
rasmussen2014phylodynamic}, our method is the first (to our knowledge) which
can fit models that do not assume panmixia. We believe this capability will be
of broad interest to the molecular evolution and epidemiology community, as it
widens the field of epidemiological parameters which may be investigated
through viral sequence data. In addition, the characterization of local contact
networks could be valuable from a public health perspective, such as for
investigating optimal vaccination strategies. This information could assist in
curtailing current epimemics, as well as preventing future epidemics of
different diseases over the same contact network.

### 2016-04-20 doc/thesis/objective.tex

\section{Objective}

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called the ``law of mass action'', implies that any two
individuals in the same compartment are equally likely to come into contact
causing transmission. Although this provides a reasonable approximation in many
cases, the error introduced by assuming a panmictic population can be
substantial when significant contact heterogeneity exists in the underlying
population. Contact network models provide an alternative to compartmental
models which do not require the assumption of panmixia. In addition to more
accurate predictions, the parameters of the networks themselves may be of
interest from a public health perspective. For example, certain vaccination
strategies may be more or less effective in curtailing an epidemic depending on
the underlying network's degree distribution. Phylodynamic methods have been
used to fit many different types of model to phylogenetic data, but as far as
we know, no methods have yet been developed to fit contact network models. The
primary objective of this work is to develop such a method.

>>>>

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact causing transmission. Although this provides a reasonable
approximation in many cases, the error introduced by assuming a panmictic
population can be substantial when significant contact heterogeneity exists in
the underlying population. Contact network models provide an alternative to
compartmental models which do not require the assumption of panmixia. In
addition to more accurate predictions, the parameters of the networks
themselves may be of interest from a public health perspective. For example,
certain vaccination strategies may be more or less effective in curtailing an
epidemic depending on the underlying network's degree distribution.
Phylodynamic methods have been used to fit many different types of model to
phylogenetic data, but as far as we know, no methods have yet been developed to
fit contact network models. The primary objective of this work is to develop
such a method.
